Linear Mixed Models: Random Intercept

### 2024-05-01 examples/3a_Examples_LinearMixedModels.qmd
new examples updates

---
title: "Linear Mixed Models: Random Intercept"
author: "Biostatistics Working Group"
execute:
  freeze: auto  # re-render only when source changes
  cache: true
---

## Overview

<p>

The LMM:ri is similar to traditional (fixed-effect) linear regression extending on this approach by including a subject-specific random-effect that allows each participant to have their own unique intercept value, in addition to the overall mean-level (fixed-effect) intercept value zzzzt

In this example, we will use the LMM:ri to analyze trajectories of height obtained across multiple measurement occasions in a sample of youth taking part in the ABCD Study. Our primary aim is to characterize stability and change in height assessments, while accounting for observations that are clustered within youth over time. To do so, we will use the LMM:ri to simultaneously model an overall sample mean trajectory (fixed effect) and subject-specific (random) effects that vary randomly about the sample mean trajectory.

</p>

## Preliminary Setup

::: panel-tabset
### Install Packages {.tabset .tabset-fade .tabset-pills}

::: blue
> **This code installs the r packages necessary for this example, if
> they are not already installed**

```{r pckg-install}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

# Create a list of required packages
packages_required <- c("tidyverse","easystats","lme4","gtsummary","report","broom","gridExtra")

# Check which packages are not installed and install them
packages_to_install <- setdiff(packages_required, rownames(installed.packages()))
if (length(packages_to_install) > 0) {
    install.packages(packages_to_install)
}

# Load the required packages
lapply(packages_required, library, character.only = TRUE)

```
:::

### Load Packages

::: blue
**This code loads the r libraries necessary for this example**

```{r lib-load}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

library(tidyverse)    # Collection of R packages for data science
library(easystats)    # Collection of R packages statistical modeling, visualization, and reporting
library(lme4)         # Linear mixed-effects models
library(gtsummary)    # Publication-ready tables
library(report)       # Easy reporting of regression analyses
library(broom)        # Tidy and augment statistical models output
library(gridExtra)    # Arrange multiple grid-based plots on a page

```
:::

```{r config}
#| echo: FALSE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE

### Config Options
#::: blue
#**This code configures knitr code chunk options**

knitr::opts_chunk$set(echo = T, message=F, warning=F, error=F, 
                      comment=NA, cache=T, code_folding=T,
                      R.options=list(width=220, digits = 3),
                      fig.align='center', 
                      out.width='75%', fig.asp=.75)
```
:::

## Descriptives Overview

::: panel-tabset
### Read and View Data {.tabset .tabset-fade .tabset-pills}

::: blue
**This code reads in and shows the data to be used in the current
example**

```{r read-data}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| output: FALSE
#| cache: FALSE

# Set the data paths
data_path_1 <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/demo5.0.rds"
data_path_2 <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"

# Read the data
data_demographics <- readRDS(data_path_1)
data_nonimaging <- readRDS(data_path_2)

# Subset the nonimaging data to include desired variables
selected_vars <- c("src_subject_id", "eventname", "nihtbx_totalcomp_fc", "anthroweightcalc", "anthroheightcalc")
subset_data <- data_nonimaging[, selected_vars]

library(dplyr)
# # Merge the datasets on 'src_subject_id' and 'eventname'
merged_data <- data_demographics %>%
  full_join(subset_data, by = c("src_subject_id", "eventname"))

# Inspect the merged data structure
str(merged_data)

# Define event names to be retained in the analysis and convert variables to appropriate data types
eventnames_to_include <- c("baseline_year_1_arm_1",
                           "1_year_follow_up_y_arm_1",
                           "2_year_follow_up_y_arm_1",
                           "3_year_follow_up_y_arm_1",
                           "4_year_follow_up_y_arm_1")

df <- merged_data %>%
  filter(eventname %in% eventnames_to_include) %>%
  mutate(
    src_subject_id = as.factor(src_subject_id),
    eventname = factor(eventname, levels = eventnames_to_include, ordered = TRUE),
    age = as.numeric(age),
    sex = as.factor(sex),
    race.4level = as.factor(race.4level),
    hisp = as.factor(hisp),
    high.educ.bl = as.factor(high.educ.bl),
    household.income.bl = as.factor(household.income.bl),
    acs_raked_propensity_score = as.numeric(acs_raked_propensity_score),
    rel_family_id.bl = as.factor(rel_family_id.bl),
    site_id_l = as.factor(site_id_l),
    nihtbx_totalcomp_fc = as.numeric(nihtbx_totalcomp_fc),
    anthroweightcalc = as.numeric(anthroweightcalc),
    anthroheightcalc = as.numeric(anthroheightcalc)
  ) %>%
  # Exclude cases from unused assessment waves
  filter(!is.na(eventname))

```
:::

### Descriptives

::: blue
**This code creates a descriptives table**

```{r descriptives}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE

descriptives_table <- df %>%
  select(eventname, sex, race.4level, hisp, anthroweightcalc) %>%
  mutate(eventname = factor(eventname, labels = c("Baseline", "Year 1","Year 2","Year 3","Year 4"))) %>%
  mutate(sex = factor(sex, labels = c("Female", "Male"))) %>%
  tbl_summary(
    by = eventname,
    missing = "no",
    label = list(sex ~ "Sex", race.4level ~ "Race", hisp ~ "Hispanic", 
                 anthroweightcalc ~ "Weight"),
    statistic = list(all_continuous() ~ "{mean} ({sd}) )", all_categorical() ~ "{p}%"),
  ) %>%
modify_header(all_stat_cols() ~ "**{level}**<br>N = {n}") %>%
  bold_labels() %>%
  italicize_levels() %>%
  modify_spanning_header(all_stat_cols() ~ "**Assessment Wave**")
theme_gtsummary_compact()

descriptives_table
```
:::
:::

## Results

::: panel-tabset
### Build Model {.tabset .tabset-fade .tabset-pills}

::: blue
The code fits a linear mixed model to predict the 'Height' variable based on time points ('eventname') and handedness ('Handedness'), while accounting for individual-level variability by including random intercepts for each participant ('src_subject_id'). The results of the model are then printed to provide a summary of the fitted model parameters.

**STEP 1: Compute LMM Model with Random Intercepts**
```{r build-model-1}
#| class.source: 'fold-hide'
#| message: FALSE
#| warning: FALSE
#| echo: TRUE
#| cache: true

## Linear Mixed Model with a random intercept (LMM-ri)
model <- lmer(anthroweightcalc ~ 1 + eventname + sex + (1|src_subject_id), data = df, REML=T)

@@

```{r model-output}
#| echo: TRUE
#| messages: FALSE
#| warning: FALSE
#| code-summary: testing

## Output and reports extending from the LMM-ri analyses
#summary(model)
#confint(model, level = 0.95, method = "Wald")

model_performance(model)

model_parameters(model)

model_parameters(model,standardize = "refit")

@@

The code provided executes a linear mixed model (LMM) to predict children's 
height across different time points (Baseline, Year_1, Year_2, Year_3, and Year_4), and also takes into account their sex. This model accounts for individual variability in height by including a random intercept for each subject (src_subject_id). The output indicates that the model was fit using the REML (Restricted Maximum Likelihood) criterion.

:::

### Model Plots

::: blue
The following set of plots are used to facilitate model diagnostics. The first is a histogram showcasing the distribution of random intercepts for individual subjects, indicating variations in height not explained by the fixed effects. The second depicts residuals versus fitted values, helping assess the model's fit and potential heteroscedasticity. The third contrasts observed and predicted height values across different time points, offering a side-by-side evaluation of the model's predictions against actual observations.

```{r diagnosticplots}

# Extract the random effects
random_effects <- ranef(model)[[1]]

# Convert to dataframe
random_effects_df <- data.frame(Intercept = random_effects$`(Intercept)`)

# Plot 1: Histogram
hist_plot <- ggplot(random_effects_df, aes(x = Intercept)) +
  geom_histogram(aes(y = ..density..), bins = 30, color = "black", fill = "lightblue") + labs(title = "Histogram of Random Effects", x = "Random Intercept Values", y = "Density") +
  theme_minimal()

# Plot 2: Residuals vs Fitted Values
resid_plot <- ggplot(NULL, aes(x = fitted_values, y = residuals)) +
  geom_point(alpha = 0.5) +
  geom_hline(yintercept = 0, linetype = "dashed") +
  labs(title = "Residuals vs Fitted Values", x = "Fitted Values", y = "Residuals") +
  theme_minimal()

@@

library(ggplot2)
theme_set(theme_modern(base_size = 6))

plot(parameters(model)) +
  ggplot2::labs(title = "A Dot-and-Whisker Plot")
#plot(model)

result <- estimate_expectation(model, data = "grid")
plot(result)
#plot(result, type = "qq")
#plot(result, type = "pp")

normality <- check_normality(model, effects = "random")
#plot(normality)

heteroscedasticity <- check_heteroscedasticity(model)
#plot(heteroscedasticity)

homogeneity <- check_homogeneity(model)
#plot(homogeneity, type = "density")

distribution <- check_distribution(model)
#plot(distribution)

#contrasts <- estimate_contrasts(model, contrast = "sex")
#means <- estimate_means(model)
#plot(means)
#plot(contrasts, means)

params <- model_parameters(model, effects = "fixed")
#plot(params, show_labels = TRUE, size_text = 4)

#check_model(model)

@@

:::

```{r trajectoryplot}

# Extract the data frame used in the model
model_data <- model@frame

# Extract unique subject IDs from the model's data
original_subject_ids <- unique(model_data$src_subject_id)

# Subset the original data to include only those subjects
df_subset <- df %>% filter(src_subject_id %in% original_subject_ids)

eventname_map <- c(
  "baseline_year_1_arm_1" = "Baseline",
  "1_year_follow_up_y_arm_1" = "Year_1",
  "2_year_follow_up_y_arm_1" = "Year_2",
  "3_year_follow_up_y_arm_1" = "Year_3",
  "4_year_follow_up_y_arm_1" = "Year_4"
)

# Apply the recoding to the eventname variable
df_subset$eventname <- factor(df_subset$eventname, levels = names(eventname_map), labels = eventname_map)

# Verify the recoding
table(df_subset$eventname)

# Generate the plot
ggplot(df_subset, aes(x = eventname, y = anthroheightcalc, group = src_subject_id)) +
  # Individual estimated height trajectories in faded lines
  geom_line(aes(group = src_subject_id), alpha = 0.3, color = "grey50") +
  # Overall group-mean trajectory in blue with increased thickness
  stat_summary(aes(group = 1), fun = mean, geom = "line", color = "blue", linewidth = 1) +
  labs(title = "Individual and Group-Mean Height Trajectories",
       x = "Event Name",
       y = "Height") +
  theme_minimal()

```{r}
#| echo: false

## Creating variable (names) from output to be used for inline text

library(broom.mixed)
library(easystats)

tidyfit <- tidy(model)
tidyaugment <- augment(model)
tidyglance <- glance(model)

Intercept_estimate <- round(tidyfit[tidyfit$term == "(Intercept)", "estimate"], 3)
Intercept_SE <- round(tidyfit[tidyfit$term == "(Intercept)", "std.error"], 3)

sex_effect <- round(tidyfit[tidyfit$term == "sexMale or other", "estimate"], 3)

# Extract values from the tibble for inline text
Intercept_estimate <- Intercept_estimate$estimate
Intercept_SE <- Intercept_SE$std.error
sex_effect <-sex_effect$estimate

#############

#r eventname_estimate
#r sex_effect
#r conditional_R2
#r conditional_R2*100**%
#r marginal_R2

```{r}
report(model)
```
:::

## Wrapping Up
<div class = "blue">

::: panel-tabset
### Write-up {.tabset .tabset-fade .tabset-pills}

::: blue

The estimate of the random intercept was `r Intercept_estimate` (*SE* `r Intercept_SE`). The fixed effects section provides the coefficients for the intercept, eventname, and the sex variable.

The linear mixed model analysis was conducted to predict children's height across different time points (Baseline, Year_1, Year_2, Year_3, and Year_4) using the event name (eventname) and height (Height) variables. The intercept, corresponding to the reference levels of the predictors is estimated at `r Intercept_estimate` units, and this effect is highly significant. Regarding sex, boys show an increase in height of about `r sex_effect` (*SE* `r Intercept_SE`) units compared to girls.

```{r}
report_performance(model)
```

:::
:::

>>>>

---
title: "Linear Mixed Models: Random Intercept"
author: "Biostatistics Working Group"
---

```{r}
#| echo: FALSE # The content of this code block will not be shown when rendered
#| output: FALSE # Output from this code block will not be shown when rendered

# This code configures knitr code chunk options

knitr::opts_chunk$set(
    echo = T, message = F, warning = F, error = F, output = T,
    comment = NA, cache = T, code_folding = T,
    R.options = list(width = 220, digits = 3),
    fig.align = "center",
    out.width = "75%", fig.asp = .75
)
```

```{r}
#| echo: FALSE # The content of this code block will not be shown when rendered
#| output: FALSE # Output from this code block will not be shown when rendered
#| cache: FALSE

# This code loads the data to be used for the example

library(tidyverse)

# Set the data path
data_path <- "/Users/shawes/ABCD/data/rds/abcd_5.0_rds/core-rds-5.0/non-imaging_excluding_nt_5.0.rds"

# Read the data
df <- readRDS(data_path)

# Create and mutate the dataframe
df_long <- df %>%
    select(src_subject_id, eventname, anthroheightcalc) %>%
    filter(eventname %in% c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1")) %>%
    drop_na(anthroheightcalc) %>%
    arrange(src_subject_id, eventname) %>%
    mutate(
        src_subject_id = factor(src_subject_id),
        eventname = factor(eventname, levels = c("baseline_year_1_arm_1", "1_year_follow_up_y_arm_1", "2_year_follow_up_y_arm_1"), labels = c("Baseline", "Year_1", "Year_2")),
        anthroheightcalc = round(as.numeric(anthroheightcalc), 2)
    ) %>%
    rename(
        id = src_subject_id,
        event = eventname,
        height = anthroheightcalc
    )

@@

## Overview
The Linear mixed model (LMM) with a random intercept is similar to traditional (fixed-effect) linear regression, extending on this approach by including a subject-specific random-effect that allows each participant to have their own unique intercept value, in addition to the overall mean-level (fixed-effect) intercept value.

This example will examine trajectories of height obtained across 3 annual measurement occasions in a sample of youth taking part in the ABCD Study. The primary aim is to characterize stability and change in participant's height, while accounting for observations that are clustered within youth over time. The random-intercept LMM will simultaneously estimate an overall sample mean (fixed effect) for the height trajectory's intercept and slope values, as well as subject-specific (random) effects that vary randomly about the mean intercept value.

## Preliminary Setup
### Load Packages
```{r}

library(tidyverse) # Collection of R packages for data science
library(gtsummary) # Publication-ready tables
library(rstatix) # Statistical Tests in tidy format
library(lme4) # Linear mixed-effects models
```

### Descriptives
```{r}

descriptives_table <- df_long %>%
    select(event, height) %>%
    mutate(event = factor(event)) %>%
    tbl_summary(
        by = event,
        missing = "no",
        label = list(height ~ "Height"),
        statistic = list(all_continuous() ~ "{mean} ({sd})", all_categorical() ~ "{p}%"),
    ) %>%
    modify_header(all_stat_cols() ~ "**{level}**<br>N = {n}") %>%
    bold_labels() %>%
    italicize_levels() %>%
    modify_spanning_header(all_stat_cols() ~ "**Assessment Wave**")
theme_gtsummary_compact()

descriptives_table

@@

## Results
### Compute LMM Model with Random Intercepts
This code fits a radom-intercept LMM to predict 'Height' across 3 annual visits, accounting for individual-level variability (random effects) of intercept values across participants. Model output is provided below, as well as a brief summary of results.

This model accounts for individual variability in height by including a random intercept for each subject (id). The output indicates that the model was fit using the REML (Restricted Maximum Likelihood) criterion.

model <- lmer(height ~ 1 + event + (1 | id), data = df_long, REML = T)

## Output and summary statistics
summary(model)
confint(model, level = 0.95, method = "Wald")

@@

These results show an increasing trajectory of height across the years, with an increase of approximately 2.36 inches in Year 1 and 5.02 inches by Year 2 compared to the baseline, as evidenced by the fixed effects estimates. The random intercepts effect suggests some between-subject variability in particpant's height values at baseline, with a standard deviation of about 3.03 inches. The model's residuals are mostly well-behaved, although the range suggests a few extreme values or outliers.

### Model Plots
```{r}

# Extract fixed effects
fixed_effects <- fixef(model)

# Prepare data frame for plotting
df_long$predicted_height <- predict(model, re.form = NA) # Overall fixed effect predictions

# Calculate individual intercept adjustments
df_long$individual_intercept <- predict(model, re.form = ~ (1 | id)) - fixed_effects[1]

# Generate the plot
ggplot(df_long, aes(x = event, y = height)) +
    geom_line(aes(y = height, color = "Individual Trajectories"), alpha = 0.3) +
    geom_line(aes(y = predicted_height + individual_intercept, group = id, color = "Adjusted Individual Trajectories"), alpha = 0.7) +
    geom_line(aes(x = event, y = predicted_height, group = 1, color = "Sample Mean Trajectory"), size = 1.5) +
    scale_color_manual(values = c("Individual Trajectories" = "red", "Adjusted Individual Trajectories" = "grey", "Sample Mean Trajectory" = "blue")) +
    labs(
        title = "Individual Trajectories with Fixed Slope and Random Intercepts",
        x = "Event Name",
        y = "Height (inches)",
        color = "Trajectory Type"
    ) +
    theme_minimal() +
    theme(legend.position = "bottom")

@@

This plot illustrates the change in participant's height across 3 annual measurement occassions. The individual raw data points are shown as red lines, representing the original height measurements, which demonstrate the variability among participants at each time point. These are overlaid with grey lines, which adjust each individual's trajectory to start from their unique intercept while following a common growth pattern indicated by the fixed effects from the model. The blue line represents the sample mean trajectory, calculated from the model's fixed effects, showcasing the average growth trend across all participants.

The plot shows individual trajectories that vary around the model's predicted values due to random effects, indicating substantial inter-individual variability in baseline heights. On average, participants experienced an increase in height of approximately 2.36 inches from baseline to Year 1 and an additional 2.66 inches from Year 1 to Year 2. The random effects for individual intercepts suggest differences in starting heights, with a standard deviation of 3.03 inches, reflecting the variability in growth patterns among the participants.

## Wrapping Up
The random-intercept LMM analysis revealed an increasing pattern of change in children's height measured over 3 annual assessments. The mean intercept value showed an average initial height of 55.24 inches (sd = 3.03) with evidence of individual variability around this (random effect) estimate. An examination of the slope factor revealed year-to-year growth increments of approxmiately 2.5 inches, with the rate of change held constant across all participants.

The LMM framework used in this example offered the ability to account for individual differences in initial height and the consistent growth patterns over time, demonstrating its utility in handling clustered or correlated data that arise from repeated measurements on the same subjects. More generally, this approach is often particularly useful in developmental studies as it allows for the inclusion of both fixed effects, which estimate common parameters across all individuals, and random effects, which capture unique variations among individuals. This enables a more precise understanding of both the group-level trends and individual-specific trajectories in longitudinal data.

### 2024-05-01 examples/examples.qmd
new examples updates

						<a href="./_9_Examples_LatentChangeScoresModels.qmd"><i class="fas fa-play

@@

						<a href="./_10_Examples_LatentGrowthCurveModels.qmd"><i class="fas fa-play

@@

>>>>

						<a href="./9_Examples_LatentChangeScoresModels.qmd"><i class="fas fa-play

@@

						<a href="./10_Examples_LatentGrowthCurveModels.qmd"><i class="fas fa-play

@@

						<a href="./13b_Examples_LatentClassGrowthAnalysis.qmd"><i class="fas fa-play
						mr-2"></i>Latent Class Growth Analysis</a>
